Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-09 analysis/paper/paper.Rmd
added petrography data and analysis

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. We identified these phases based on 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`pre_chi`$) or post-European periods (p= $`post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

>>>>

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. These phases were identified based using 32 previously published radiocarbon ages [@Chen2007], and enlisted excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.

@@

We used scans of pottery drawings at 300 dpi acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].
